---
title: "Practicum I CS5200"
output: html_notebook
author: "Pranay Shukla and Deepthi Varanasi"
date: "Fall 2023"
---

#Installing all required packages
```{r addingPackages, eval=T}
library(RMySQL)
library(sqldf)
options(sqldf.driver = "SQLite")
library(dplyr)
library(ggplot2)
```

## connect to the database
```{r connectingToDatabase, eval=T }
dbcon <- dbConnect(MySQL(), 
                user = "sql9656433", 
                password = "D3pzfkEWXg", 
                dbname = "sql9656433", 
                host = "sql9.freemysqlhosting.net")
```

# Checking the databases present
```{sql showDatabase, connection=dbcon}
SHOW DATABASES

```

# Using the DB 
```{sql usingDatabase, connection=dbcon}
USE sql9656433
```
# Dropping tables if already exists

```{sql dropTableIfExist, connection=dbcon}
DROP TABLE IF EXISTS airports
```

```{sql dropTableIfExist, connection=dbcon}
DROP TABLE IF EXISTS flights
```

```{sql dropTableIfExists, connection=dbcon}
DROP TABLE IF EXISTS conditions
```

```{sql dropTableIfExists, connection=dbcon}
DROP TABLE IF EXISTS strikes
```



# Query - 1
# Creating table flights and checking altitude has only positive values

```{sql createFligts, connection=dbcon}
CREATE TABLE flights (
    fid INTEGER PRIMARY KEY,
    date DATE,
    origin INTEGER,
    airline TEXT,
    aircraft TEXT,
    altitude INTEGER CHECK (altitude >= 0),
    heavy BOOLEAN
);

```

# creating table airports

```{sql createAirport, connection=dbcon}
CREATE TABLE airports(
aid INTEGER NOT NULL AUTO_INCREMENT, 
airportName TEXT NOT NULL, 
airportCode TEXT , 
airportState TEXT NOT NULL,
PRIMARY KEY(aid)
)

```

# customising flights to have foreign key

```{sql alterFlight, connection=dbcon}
ALTER TABLE flights
ADD CONSTRAINT fk_origin_airports
FOREIGN KEY (origin) REFERENCES airports (aid);
```

# Creating a lookup table conditions which stores the cid, condition and explanation

```{sql lookUpTb, connection=dbcon}
CREATE TABLE conditions(
cid INTEGER NOT NULL AUTO_INCREMENT, 
`sky_condition` TEXT NOT NULL, 
explanation TEXT,
PRIMARY KEY(cid)
)
```



# creating strikes
# part f included adding fid as the foreign key
```{sql strikesTabl, connection=dbcon}
CREATE TABLE strikes (
    sid INTEGER PRIMARY KEY,
    fid INTEGER,
    numbirds INTEGER,
    impact TEXT,
    damage BOOLEAN,
    altitude INTEGER CHECK (altitude >= 0),
    conditions INTEGER,
    FOREIGN KEY (fid) REFERENCES flights (fid),
    FOREIGN KEY (conditions) REFERENCES conditions (cid)
);
```


# code to test table definitions







# Loading the CSV file into R without filepath

```{r loadingCSV, eval = T}
fn <- "BirdStrikesData-V2.csv"
bds.raw <- read.csv(fn, header = TRUE, stringsAsFactors = FALSE)
```

# checking if we got the csvFile
```{r printingCSV, eval = T}
bds.raw
```

# we store Caused damage as true and No damage as false

```{r changeCauseDamageToTrueAndViceVersa, eval = T}
bds.raw$damage <- ifelse(bds.raw$damage == 'Caused damage', TRUE, FALSE)
bds.raw
```

#populating the empty fields with unknown..
 
```{r populatingTheEmptyRows, eval = T}
options(warn = -1) 
bds.raw$airline[bds.raw$airline == ""] <- "Unknown"
```

```{r}
options(warn = -1) 
bds.raw$airline[bds.raw$airport == ""] <- "Unknown"
```

# Creating a dataframe for conditions
```{r dataframeCond, eval = T}
cond_df <-data.frame(sky_condition= unique(bds.raw$sky_conditions))
cond_df <- cbind(cid = 1:nrow(cond_df), cond_df) 
cond_df
```
# Creating a dataframe for airports

```{r airportDataframe, eval = T}
airport_df <- data.frame(airportName=bds.raw$airport,airportState=bds.raw$origin)
#removing duplicates
airport_df <- airport_df %>% distinct()
airport_df <- cbind(aid = 1:nrow(airport_df), airport_df) 
airport_df
```
# Creating a strikes dataframe
```{r dataframeStrikes, eval = T}
strikes_df <- data.frame(`rid` = bds.raw$rid,
  numbirds = bds.raw$wildlife_struck,
                           impact = bds.raw$impact,
                           damage=bds.raw$damage,
                           altitude = bds.raw$altitude_ft, 
                           conditions = bds.raw$sky_conditions,
                         fid = 0)
strikes_df <- cbind(sid = 1:nrow(strikes_df), strikes_df) 
strikes_df
```
# creating the flights dataframe, we use rid to reference data in airports to 
# populate the field aid
```{r flightsDataframe, eval = TRUE}
flights_df <- data.frame(`rid` = bds.raw$rid,
  date = bds.raw$flight_date,
                           origin = bds.raw$origin,
  airport = bds.raw$airport,
                           airline=bds.raw$airline,
                           aircraft = bds.raw$aircraft, 
                           altitude = bds.raw$altitude_ft,
                           heavy = bds.raw$heavy_flag)
flights_df <- cbind(fid = 1:nrow(flights_df), flights_df) 
```
```{r}
flights_df
```
#converting the date to datetime object

```{r}
flights_df$`date` <- as.Date(flights_df$`date`, format = "%m/%d/%Y")
flights_df
```

# Checking null values for each column
```{r}

sapply(flights_df, function(x) sum(is.na(x)))

```




# Omitting the null values from flights dataframe
```{r}
flights_df <- na.omit(flights_df) 
flights_df
```

# Checking again for null values in each column
```{r}

sapply(flights_df, function(x) sum(is.na(x)))

```


# Replacing the origin in flights table with unique id (aid) from airports dataframe
```{r}
flights_df <- sqldf("SELECT fid,flights_df.rid,`date`,airline,aircraft, altitude, heavy, (SELECT aid FROM airport_df WHERE airportState = origin AND airportName = flights_df.airport ) AS origin FROM flights_df")
                   


```

```{r}
flights_df
```


# changing fid in strikes to flights' fid
```{r}
strikes_df <- sqldf("
  SELECT s.sid, s.rid, f.fid, s.numbirds, s.impact, s.damage, s.altitude, s.conditions
  FROM strikes_df s
  LEFT JOIN flights_df f ON s.rid = f.rid
")
```

```{r}
strikes_df
```



# Writing the airports dataframe to airports table


```{r}
dbWriteTable(dbcon, "airports", airport_df[1:nrow(airport_df),], row.names = FALSE, append = TRUE)
```



#writing conditions df to conditions

```{r}
dbWriteTable(dbcon, "conditions", cond_df[1:nrow(cond_df),], row.names = FALSE, append = TRUE)
```




# replacing conditions with condition id in strikes df

```{r}
strikes_df <- sqldf("SELECT `rid`,sid,fid,numbirds,impact,damage,altitude,cid
AS conditions FROM strikes_df
JOIN cond_df ON strikes_df.conditions = cond_df.sky_condition")
strikes_df
```



# removing rid from strikes

```{r}
strikes_df <- sqldf("SELECT s.sid, s.fid, s.numbirds, s.impact, s.damage, s.altitude, s.conditions
  FROM strikes_df s")
```

```{r}
strikes_df
```




# populating the strikes table:- 
```{r papulatingTheStrikesTable, eval = T}
dbWriteTable(dbcon, "strikes", strikes_df[1:nrow(strikes_df),], row.names = FALSE, append = TRUE)
```

# removing rid from flights

```{r}
flights_df <- sqldf("SELECT fid,`date`,airline,aircraft, altitude, heavy, origin from flights_df")
```
```{r}
flights_df
```
#populating the flights table

```{r}
dbWriteTable(dbcon, "flights", flights_df[1:nrow(flights_df),], row.names = FALSE, append = TRUE)
```



# checking if the data has been populated properly

```{sql connection = dbcon}
SELECT * from conditions LIMIT 2
```

```{sql connection = dbcon}
SELECT * from airports LIMIT 19
```
```{sql connection = dbcon}
SELECT * from flights LIMIT 7
```

# populating the strikes table:- 
```{r papulatingTheStrikesTable, eval = T}
dbWriteTable(dbcon, "strikes", strikes_df[1:nrow(strikes_df),], row.names = FALSE, append = TRUE)
```

```{sql connection = dbcon}
SELECT * from strikes LIMIT 17
```
#top 10 states with maximum bird incidents 

```{sql connection=dbcon}
SELECT a.airportState, COUNT(s.sid) AS num_incidents
FROM strikes AS s
JOIN flights AS f ON s.fid = f.fid
JOIN airports AS a ON f.origin = a.aid
GROUP BY a.airportState
ORDER BY num_incidents DESC
LIMIT 10;

```
#9 Above average number of bird strikes

```{sql connection=dbcon}
SELECT f.airline, COUNT(s.sid) AS num_incidents
FROM strikes AS s
JOIN flights AS f ON s.fid = f.fid
GROUP BY f.airline
HAVING num_incidents > (SELECT AVG(num_incidents) FROM (SELECT f.airline, COUNT(s.sid) AS num_incidents FROM strikes AS s JOIN flights AS f ON s.fid = f.fid GROUP BY f.airline) AS avg_table);

```

# monthly bird strike
```{sql connection=dbcon}
SELECT DATE_FORMAT(`date`, '%m') AS month, SUM(numbirds) AS total_birds
FROM strikes AS s
JOIN flights AS f ON s.fid = f.fid
GROUP BY month
LIMIT 6
```
#plot graph q 11
```{r}
# Execute your SQL query and store the result in a variable
query <- "SELECT DATE_FORMAT(`date`, '%m') AS month, SUM(numbirds) AS total_birds
FROM strikes AS s
JOIN flights AS f ON s.fid = f.fid
GROUP BY month"
rs <- dbGetQuery(dbcon, query)

# Create a line plot
plot(rs$month, rs$total_birds, type = "l", 
     xlab = "Month", ylab = "Number of Birds",
     main = "Number of Birds vs. Month")

# Add a title, x-axis label, and y-axis label
title(main = "Number of Birds vs. Month", sub = "", xlab = "Month", ylab = "Number of Birds")

```






# Q 12 stored procedure
# Creating a stored procedure that adds a new incident to the database

```{sql connection=dbcon}
DROP PROCEDURE IF EXISTS AddBirdStrike
```

```{sql connection=dbcon}

CREATE PROCEDURE AddBirdStrike(
    IN p_date DATE,
    IN p_airline TEXT,
    IN p_aircraft TEXT,
    IN p_altitude INTEGER,
    IN p_heavy BOOLEAN,
    IN p_numbirds INTEGER,
    IN p_impact TEXT,
    IN p_damage BOOLEAN,
    IN p_sky_condition TEXT,
    IN p_airport_name TEXT,
    IN p_airport_code TEXT,
    IN p_airport_state TEXT
)
BEGIN
    DECLARE v_flight_id INT;
    DECLARE v_condition_id INT;
    DECLARE v_airport_id INT;

    -- Insert or update the flight record
    INSERT INTO flights (`date`, airline, aircraft, altitude, heavy)
    VALUES (p_date, p_airline, p_aircraft, p_altitude, p_heavy)
    ON DUPLICATE KEY UPDATE fid = LAST_INSERT_ID(fid);

    -- Get the flight ID
    SELECT LAST_INSERT_ID() INTO v_flight_id;

    -- Insert or update the condition record
    INSERT INTO conditions (sky_condition)
    VALUES (p_sky_condition)
    ON DUPLICATE KEY UPDATE cid = LAST_INSERT_ID(cid);

    -- Get the condition ID
    SELECT LAST_INSERT_ID() INTO v_condition_id;

    -- Insert or update the airport record
    INSERT INTO airports (airportName, airportCode, airportState)
    VALUES (p_airport_name, p_airport_code, p_airport_state)
    ON DUPLICATE KEY UPDATE aid = LAST_INSERT_ID(aid);

    -- Get the airport ID
    SELECT LAST_INSERT_ID() INTO v_airport_id;

    -- Insert the new bird strike incident
    INSERT INTO strikes (fid, numbirds, impact, damage, altitude, conditions)
    VALUES (v_flight_id, p_numbirds, p_impact, p_damage, p_altitude, v_condition_id);

    -- You can add more code here if needed

END;

```
```{sql connection = dbcon}
CALL AddBirdStrike("2023-08-30","lufthansa","FlyUS",10,TRUE,10,"full impact",TRUE,"Sunny","Chaudhry Charan", "123123", "UP")
```
# Procedure ran, Sunny is inserted later in the table

```{sql connection=dbcon}
SELECT * FROM conditions
```



